Making ice sheet models \\ scale properly

### 2023-02-10 talk/slides.tex
I think this is good enough

\bigskip
\begin{center}
\begin{minipage}{0.82\textwidth}
how does this metric \alert{scale} in the \alert{high spatial resolution limit} $h\to 0$, equivalently $m\to \infty$?
\end{minipage}
\end{center}
\end{itemize}
\end{frame}

\begin{frame}{\emph{explicit} ice sheet model performance}

\begin{tabular}{llll}
\emph{time-stepping} &  & \emph{flops per model year} \\ \hline
\\
explicit & SIA    & $\oo{\frac{D\, L^2}{h^{\,4}}} = \oo{\frac{D}{L^2} m^{\,2}}$ \\
\\
explicit ({\footnotesize \emph{advective}}) & Stokes & $\oo{\frac{U \,L^{2+2\alpha}}{h^{\,3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
\\
\phantom{explicit} ({\footnotesize \emph{diffusive}})  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{\,4+2\alpha}}} = \oo{\frac{D}{L^2} m^{\,2+\alpha}}$
\end{tabular}

\bigskip
\begin{itemize}
\item \alert{explicit} time-stepping \alert{implies many stress-balance solves}
    \begin{itemize}
    \item[$\circ$] stress-balance solver scaling $\alpha$ is important,
    \item[$\circ$] but even optimality ($\alpha=0$) cannot rescue performance
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item switch to \alert{implicit time-stepping} for unconditional stability?
    \begin{itemize}
    \item[$\circ$] each step is a \alert{free-boundary} NCP-coupled-to-Stokes problem
    \item[$\circ$] parameterize cost of these solves as $O(m^{1+\beta})$
    \end{itemize}
\item nonetheless need $q$ model updates per year to integrate climate influences
\end{itemize}
\end{frame}

@@

\begin{tabular}{llll}
\emph{time-stepping} &  & \emph{flops per model year} \\ \hline
\\
explicit & SIA    & $\oo{\frac{D\, L^2}{h^{\,4}}} = \oo{\frac{D}{L^2} m^{\,2}}$ \\
\\
explicit ({\footnotesize \emph{advective}}) & Stokes & $\oo{\frac{U \,L^{2+2\alpha}}{h^{\,3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
\\
\phantom{explicit} ({\footnotesize \emph{diffusive}})  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{\,4+2\alpha}}} = \oo{\frac{D}{L^2} m^{\,2+\alpha}}$ \\
\\
implicit & \emph{either} & $\oo{\frac{q\, L^{2+2\beta}}{h^{\,2+2\beta}}} = \oo{q\, m^{1+\beta}}$
\end{tabular}

\bigskip
\begin{itemize}
\item to really reduce power on $m$, use \alert{implicit} time-stepping and build a \alert{near-optimal ($\beta \approx 0$) NCP-coupled-to-Stokes solver}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item no convincing NCP-coupled-to-Stokes (free-boundary) solvers exist yet
    \begin{itemize}
    \item[$\circ$] Wirbel \& Jarosch~(2020) is an important attempt \dots
    \end{itemize}
\item the Bueler (2016) free-boundary solver uses the SIA
    \begin{itemize}
    \item[$\circ$] and it scales badly: $\beta=0.8$
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item idea from L{\"o}fgren et al.~(2022)
    \begin{itemize}
    \item[$\circ$] earlier use in mantle/crust simulations (Kaus et al.~2010)
    \end{itemize}
\item \emph{idea.} keep time-splitting (de-coupled Stokes, SKE solves) but \alert{modify the Stokes problem to see the updated surface}
\item that is, modify the Stokes weak form \onslide<2>{by adding body force terms corresponding to the updated-surface icy domain}

\vspace{-3mm}
\begin{minipage}[t][30mm][t]{0.8\textwidth}
\only<1>{
{\small
\begin{align*}
\int_{\Lambda^\ell} 2 \nu D\bu : D\bv - \int_{\Lambda^\ell} p \Div \bv &= \int_{\Lambda^\ell} \bbf\cdot \bv \\
\int_{\Lambda^\ell} q \Div \bu &= 0
\end{align*}
}
}
\only<2>{
{\small
\begin{align*}
\int_{\Lambda^\ell} 2 \nu D\bu : D\bv - \int_{\Lambda^\ell} p \Div \bv \qquad\qquad &= \int_{\Lambda^\ell} \bbf\cdot \bv + \Delta t \int_{\Gamma_s^\ell} a\, (\bbf\cdot \bv)\,dx\\
\hspace{20mm} - \Delta t \int_{\Gamma_s^\ell} (\bu \cdot \bn_s^\ell) (\bbf\cdot \bv)\,dx \qquad
&\phantom{=} \\
\int_{\Lambda^\ell} q \Div \bu &= 0
\end{align*}
}
}
\end{minipage}

@@

\begin{itemize}
\item direct attack on the problem seems to require a \alert{multilevel} solver for \alert{variational inequalities} (VIs), but in the \alert{non-local residual case}
    \begin{itemize}
    \item[$\circ$] this seems not to exist
    \item[$\circ$] the \alert{smoother} must compute an implicit-step SKE residual including surface-motion term $\Phi(s) = - \bu|_s\cdot \bn_s$ from a scalable Stokes solver
    \end{itemize}
\item near-optimal multilevel VI solvers exist for toy problems
    \begin{itemize}
    \item[$\circ$] Poisson equation obstacle problem (Brandt \& Cryer 1983; Gr\"aeser \& Kornhuber 2009)
    \end{itemize}
\item<2> join me on this quest?
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item ice sheet models solve a multi-scale, irregular-data problem with hard-to-observe boundary conditions
   \begin{itemize}
   \item[$\circ$] there are \alert{no easy or magic techniques} for performance
   \end{itemize}
\item<2-> current-technology ice sheet models mostly use \alert{explicit} time stepping, \alert{non-optimal} stress-balance solvers, and \alert{shallow} assumptions
   \begin{itemize}
   \item[$\circ$] progress will be made in all of these areas
   \end{itemize}
\item<3-> \alert{coming soon} from current research:
   \begin{enumerate}
   \item[1.] machine learning emulators (Jouvet et al.~2021)
   \item[2.] semi-coupled time stepping (L{\"o}fgren et al.~2022)
   \item[3.] scalable Stokes solvers (Isaac et al.~2015)   
   \end{enumerate}
\item<4-> scalable solvers for implicit-step NCP-coupled-to-Stokes models require advances in \alert{multilevel solutions of non-local variational inequalities}
\end{itemize}
\end{frame}

>>>>

\bigskip
\begin{center}
\begin{minipage}{0.82\textwidth}
how does this metric \alert{scale} in the \alert{high spatial resolution limit} $h\to 0$, equivalently $m\to \infty$?
\end{minipage}
\end{center}

\bigskip
\item the goal: \hspace{20mm} $O(h^{-2}) = O(m^1)$
\end{itemize}
\end{frame}

\begin{frame}{explicit ice sheet model performance}

\begin{tabular}{llll}
\emph{time-stepping} &  & \emph{flops per model year} \\ \hline
\\
explicit & SIA    & $\oo{\frac{D\, L^2}{h^{\,4}}} = \oo{\frac{D}{L^2} m^{\,2}}$ \\
\\
explicit ({\footnotesize \emph{advective}}) & Stokes \phantom{xxxx} & $\oo{\frac{U \,L^{2+2\alpha}}{h^{\,3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
\\
\phantom{explicit} ({\footnotesize \emph{diffusive}})  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{\,4+2\alpha}}} = \oo{\frac{D}{L^2} m^{\,2+\alpha}}$
\end{tabular}

\vspace{10mm}
\begin{itemize}
\item explicit time-stepping implies \alert{many stress-balance solves}
    \begin{itemize}
    \item[$\circ$] while stress-balance scaling exponent $\alpha$ is important, even optimality ($\alpha=0$) cannot rescue performance
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item switch to \alert{implicit time-stepping} for unconditional stability?
    \begin{itemize}
    \item[$\circ$] each step is a \alert{free-boundary} NCP-coupled-to-Stokes problem
    \item[$\circ$] parameterize cost of these solves as $O(m^{1+\beta})$
    \end{itemize}
\item need $q$ model updates per year to integrate climate influences
\end{itemize}
\end{frame}

@@

\begin{tabular}{llll}
\emph{time-stepping} &  & \emph{flops per model year} \\ \hline
\\
explicit & SIA    & $\oo{\frac{D\, L^2}{h^{\,4}}} = \oo{\frac{D}{L^2} m^{\,2}}$ \\
\\
explicit ({\footnotesize \emph{advective}}) & Stokes \phantom{xxxx} & $\oo{\frac{U \,L^{2+2\alpha}}{h^{\,3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
\\
\phantom{explicit} ({\footnotesize \emph{diffusive}})  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{\,4+2\alpha}}} = \oo{\frac{D}{L^2} m^{\,2+\alpha}}$ \\
\\
implicit & & $\oo{\frac{q\, L^{2+2\beta}}{h^{\,2+2\beta}}} = \oo{q\, m^{1+\beta}}$
\end{tabular}

\bigskip
\begin{itemize}
\item goal for optimists: implicit time-stepping \emph{and} build a $\beta \approx 0$ NCP-coupled-to-Stokes solver for the time step problems
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item no convincing NCP-coupled-to-Stokes (free-boundary) solvers exist yet
    \begin{itemize}
    \item[$\circ$] Wirbel \& Jarosch~(2020) is an important attempt \dots
    \end{itemize}
\item the Bueler (2016) implicit (free-boundary) SIA solver scales badly: $\beta=0.8$
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item idea from L{\"o}fgren et al.~(2022)
    \begin{itemize}
    \item[$\circ$] earlier use in mantle/crust simulations (Kaus et al.~2010)
    \end{itemize}
\item \emph{idea.} remain explicit, but \alert{modify the Stokes problem to ``see'' the updated (extrapolated) surface}
\item that is, modify the Stokes problem at time $t^\ell$ \onslide<2>{by adding body force terms corresponding to the updated-surface icy domain}

\vspace{-3mm}
\begin{minipage}[t][30mm][t]{0.8\textwidth}
\only<1>{
{\small
\begin{align*}
\int_{\Lambda^\ell} 2 \nu D\bu : D\bv - \int_{\Lambda^\ell} p \Div \bv &= \int_{\Lambda^\ell} \bbf\cdot \bv \\
\phantom{x} \\
\phantom{x} \\
\int_{\Lambda^\ell} q \Div \bu &= 0
\end{align*}
}
}
\only<2>{
{\small
\begin{align*}
\int_{\Lambda^\ell} 2 \nu D\bu : D\bv - \int_{\Lambda^\ell} p \Div \bv \qquad\qquad &= \int_{\Lambda^\ell} \bbf\cdot \bv + \Delta t \int_{\Gamma_s^\ell} a\, (\bbf\cdot \bv)\,dx\\
\hspace{20mm} - \Delta t \int_{\Gamma_s^\ell} (\bu \cdot \bn_{s^\ell}) (\bbf\cdot \bv)\,dx \qquad
&\phantom{=} \\
\int_{\Lambda^\ell} q \Div \bu &= 0
\end{align*}
}
}
\end{minipage}

@@

\begin{itemize}
\item direct attack on the problem seems to require a \alert{multilevel} solver for \alert{variational inequalities} (VIs)
\item but in the \alert{non-local residual case} \hfill $\gets$ \emph{yesterday's seminar}
    \begin{itemize}
    \item[$\circ$] this seems not to exist
    \item[$\circ$] the \alert{smoother} must reduce a residual formed from surface-motion term $\Phi(s) = - \bu|_s\cdot \bn_s$ (from a scalable Stokes solver)
    \end{itemize}
\item near-optimal multilevel solvers exist for simpler VI problems
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item ice sheet models solve a multi-scale, irregular-data problem with hard-to-observe boundary conditions
   \begin{itemize}
   \item[$\circ$] there are \alert{no easy or magic techniques} for performance
   \end{itemize}
\item<2-> current-technology ice sheet models mostly use \alert{explicit} time stepping, \alert{non-optimal} stress-balance solvers, and \alert{shallow} assumptions
   \begin{itemize}
   \item[$\circ$] progress is being made in all of these areas
   \end{itemize}
\item<3-> \alert{coming soon} from current research:
   \begin{enumerate}
   \item[1.] machine learning emulators (Jouvet et al.~2021)
   \item[2.] semi-coupled time stepping (L{\"o}fgren et al.~2022)
   \item[3.] scalable Stokes solvers (Isaac et al.~2015)   
   \end{enumerate}
\item<4-> scalable solvers for implicit-step NCP-coupled-to-Stokes models, which would seem to be the recommended numerical design, require \alert{multilevel solvers for non-local variational inequalities}
\end{itemize}
\end{frame}
